---
title: "Actuarial Statistics Assignment"
subtitle: "Life Tables, Annuities, Loans, Survival Analysis, Multiple Decrements & Joint-Life Models"
author: "Ashik"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo
execute:
  eval: false
  warning: false
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
#| label: setup
#| include: false
library(lifecontingencies)
library(survival)
```

# Problem 1: Life Table and Actuarial Table

|  |  |  |  |  |  |  |  |  |  |  |  |
|----|----|----|----|----|----|----|----|----|----|----|----|
| ![](data:image/png;base64,iVBORw0KGgoAAAANSUhEUgAAAAcAAAAQCAYAAADagWXwAAAAAXNSR0IArs4c6QAAAJhJREFUKFOl0bEKQXEUgPHfLcpikZWy2K02ZTFaDDaPYTRalUXeAaNRWaw2ryCxoAxK/7q3uN0s96xffX3nnMifiXLDEpo4xqYCaol2iQHGWGCDZ4BlVNHGECtscf8OquCEPvZBn649o4FHGnYxwwiHBNbxQg8tXDBHMWh3uMW1HawxwTTAUBrgO94xhF2zgn6umf+2mc/5AN7qGAeIXKESAAAAAElFTkSuQmCC) | 0 | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 |
| ![](data:image/png;base64,iVBORw0KGgoAAAANSUhEUgAAAAoAAAARCAYAAADkIz3lAAAAAXNSR0IArs4c6QAAAOVJREFUKFOF0SFLhEEQxvHfgSYFo0UwmcxySZtRk1/AZhI02MRw7cADDRpF0KAgIvcBDkXBJJg0GP0EHiIYRAb3PdbjZW/L7D7735l5Zhv+1gyamEUnaf9CI50WsY97bJfAuHvAEc5LYGT+wAJeS+AcnjCFnxK4hk0s1UGhVWZamMTWKLCLC5yNAt+xjJcSOI23zEi0c5LMHaId1UJcwS7W8ZkyjuEq6X08B7iBvRRvstIBHuCucj2OyPA11N8pjvGYjydn4tE8VhFl41u/qznm4DWihR5usYPLOnAiMzXY14G1o/wFuBIpV20Q98oAAAAASUVORK5CYII=) | 900 | 850 | 800 | 750 | 700 | 650 | 600 | 550 | 500 | 450 | 400 |
| ![](data:image/png;base64,iVBORw0KGgoAAAANSUhEUgAAAAcAAAAQCAYAAADagWXwAAAAAXNSR0IArs4c6QAAAJhJREFUKFOl0bEKQXEUgPHfLcpikZWy2K02ZTFaDDaPYTRalUXeAaNRWaw2ryCxoAxK/7q3uN0s96xffX3nnMifiXLDEpo4xqYCaol2iQHGWGCDZ4BlVNHGECtscf8OquCEPvZBn649o4FHGnYxwwiHBNbxQg8tXDBHMWh3uMW1HawxwTTAUBrgO94xhF2zgn6umf+2mc/5AN7qGAeIXKESAAAAAElFTkSuQmCC) | 11 | 12 | 13 | 14 | 15 | 16 | 17 | 18 |   |   |   |
| ![](data:image/png;base64,iVBORw0KGgoAAAANSUhEUgAAAAoAAAARCAYAAADkIz3lAAAAAXNSR0IArs4c6QAAAOVJREFUKFOF0SFLhEEQxvHfgSYFo0UwmcxySZtRk1/AZhI02MRw7cADDRpF0KAgIvcBDkXBJJg0GP0EHiIYRAb3PdbjZW/L7D7735l5Zhv+1gyamEUnaf9CI50WsY97bJfAuHvAEc5LYGT+wAJeS+AcnjCFnxK4hk0s1UGhVWZamMTWKLCLC5yNAt+xjJcSOI23zEi0c5LMHaId1UJcwS7W8ZkyjuEq6X08B7iBvRRvstIBHuCucj2OyPA11N8pjvGYjydn4tE8VhFl41u/qznm4DWihR5usYPLOnAiMzXY14G1o/wFuBIpV20Q98oAAAAASUVORK5CYII=) | 350 | 300 | 250 | 200 | 150 | 100 | 50 | 0 |   |   |   |

```{r p1-setup}
#| label: p1-setup
x_sample  <- seq(from = 0, to = 18, by = 1)
lx_sample <- c(900, 850, 800, 750, 700, 650, 600, 550, 500, 450,
               400, 350, 300, 250, 200, 150, 100, 50, 0)
```

## a) Construct the life table

```{r p1a}
#| label: p1a
lifetable <- new("lifetable", x = x_sample, lx = lx_sample, name = "lifetable")
print(lifetable)
```

## b) Actuarial table at i = 3%

```{r p1b}
#| label: p1b
exampleAct <- new("actuarialtable", x = x_sample, lx = lx_sample,
                   interest = 0.03, name = "example actuarialtable")
print(exampleAct)
```

## c) Probability a life aged 5 survives 1 year: $p_{5,1}$

```{r p1c}
#| label: p1c
DemoEX1 <- pxt(lifetable, 5, 1)
DemoEX1
# p_5,1 = l6/l5 = 600/650 = 0.9231
```

## d) Probability a life aged 5 dies within 1 year: $q_{5,1}$

```{r p1d}
#| label: p1d
DemoEX2 <- qxt(lifetable, 5, 1)
DemoEX2
# q_5,1 = 1 - p_5,1 = 0.0769
```

## e) Probability a life aged 5 does NOT die between ages 6 and 7

```{r p1e}
#| label: p1e
l5 <- 650; l6 <- 600; l7 <- 550
ans <- 1 - (l6 - l7) / l5
ans
# = 1 - 50/650 = 0.9231
```

## f) Complete expectation of life at age 5

```{r p1f}
#| label: p1f
DemoEX3 <- exn(lifetable, 5, 1, "complete")
DemoEX3
# e-degree-5 = T5/l5 = 4225/650 = 6.5 years
# (matches De Moivre formula: (omega - x)/2 = (18-5)/2 = 6.5)
```

## g) Central death rate at age 5

```{r p1g}
#| label: p1g
DemoEX4 <- mxt(lifetable, 5, 1)
DemoEX4
# m5 = d5/L5 = 50/625 = 0.08
```

## h) Expected number of deaths between ages 5 and 6

```{r p1h}
#| label: p1h
DemoEX5 <- dxt(lifetable, 5, 1)
DemoEX5
# d5 = 50
```

## i) Expected total deaths between ages 10 and 15

```{r p1i}
#| label: p1i
DemoEX10 <- dxt(lifetable, 10, 5)
DemoEX10
# l10 - l15 = 400 - 150 = 250
```

## j) 15-year term insurance, age 25, i = 3%

```{r p1j}
#| label: p1j
#| eval: false
# NOT computable with this table: it only runs to age 18.
# Formula once a suitable table exists (age up to 40+):
Axn(actuarialtable = exampleAct, x = 25, n = 15)
```

## k) 10-year term insurance, age 22, i = 6%

```{r p1k}
#| label: p1k
#| eval: false
# NOT computable with this table for the same reason as (j).
Axn(actuarialtable = exampleAct, x = 22, n = 10, i = 0.06)
```

## l) Assurances on a life aged 10 (sum payable at end of year of death)

```{r p1l}
#| label: p1l
# i) Whole-life assurance of 1000
WL10 <- 1000 * Axn(actuarialtable = exampleAct, x = 10)
WL10   # ~ 877.46

# ii) Pure endowment assurance of 500, term 6 years
PE10_6 <- 500 * Exn(actuarialtable = exampleAct, x = 10, n = 6)
PE10_6 # ~ 104.69

# iii) Temporary (term) assurance of 1000, term 6 years
Term10_6 <- 1000 * Axn(actuarialtable = exampleAct, x = 10, n = 6, type = "term")
Term10_6 # ~ 677.15

# iv) Endowment assurance of 100, term 5 years
End10_5 <- 100 * Axn(actuarialtable = exampleAct, x = 10, n = 5, type = "EA")
End10_5  # ~ 89.59

# If your lifecontingencies version lacks type = "EA", build it manually:
# End10_5 <- 100 * (Axn(exampleAct, x = 10, n = 5, type = "term") +
#                    Exn(exampleAct, x = 10, n = 5))
```

# Problem 2: Loans, Savings Plans & Annuities

## a) Loan of Tk. 150,000 at 7%, 10-year term — monthly installment & amortization schedule

```{r p2a}
#| label: p2a
i <- 0.07
Capital <- 150000
n_years <- 10
n_months <- n_years * 12

# Effective monthly rate equivalent to the 7% annual rate
monthlyInt <- (1 + i)^(1/12) - 1

# Monthly installment
R <- (1/12) * Capital / annuity(i = i, n = n_years, k = 12, type = "immediate")
R   # ~ 1725.05

# Amortization schedule
balance   <- numeric(n_months + 1)
interests <- numeric(n_months + 1)
capitals  <- numeric(n_months + 1)
balance[1] <- Capital

for (m in 2:(n_months + 1)) {
  interests[m] <- balance[m - 1] * monthlyInt
  capitals[m]  <- R - interests[m]
  balance[m]   <- balance[m - 1] * (1 + monthlyInt) - R
}

loanSummary <- data.frame(month = 0:n_months,
                           payment = c(0, rep(R, n_months)),
                           interest = interests,
                           capital = capitals,
                           balance = balance)
loanSummary

# Capital and balance at month 30
loanSummary[loanSummary$month == 30, ]
```

## b) Required annual savings and annuity values for five savers

```{r p2b}
#| label: p2b
RA <- 150000 / accumulatedValue(i = 0.07,  n = 10)
RB <- 120000 / accumulatedValue(i = 0.06,  n = 12)
RC <- 200000 / accumulatedValue(i = 0.08,  n = 15)
RD <- 100000 / accumulatedValue(i = 0.075, n = 10)
RE <- 160000 / accumulatedValue(i = 0.08,  n = 20)
c(A = RA, B = RB, C = RC, D = RD, E = RE)

iannuity_A <- annuity(i = 0.07,  n = 10, k = 1,  type = "immediate")
dannuity_A <- annuity(i = 0.07,  n = 10, k = 12, type = "due")
c(iannuity_A, dannuity_A)

iannuity_B <- annuity(i = 0.06,  n = 12, k = 1,  type = "immediate")
dannuity_B <- annuity(i = 0.06,  n = 12, k = 12, type = "due")
c(iannuity_B, dannuity_B)

iannuity_C <- annuity(i = 0.08,  n = 15, k = 1,  type = "immediate")
dannuity_C <- annuity(i = 0.08,  n = 15, k = 12, type = "due")
c(iannuity_C, dannuity_C)

iannuity_D <- annuity(i = 0.075, n = 10, k = 1,  type = "immediate")
dannuity_D <- annuity(i = 0.075, n = 10, k = 12, type = "due")
c(iannuity_D, dannuity_D)

iannuity_E <- annuity(i = 0.08,  n = 20, k = 1,  type = "immediate")
dannuity_E <- annuity(i = 0.08,  n = 20, k = 12, type = "due")
c(iannuity_E, dannuity_E)
```

## c) PV of 5 annual payments of Tk. 500 at i = 4%

```{r p2c}
#| label: p2c
i <- 0.04
V <- 1 / (1 + i)

# (i) first payment at end of year 1 (annuity-immediate)
PV_immediate <- 500 * ((1 - V^5) / i)
PV_immediate   # ~ 2225.91

# (ii) first payment at time 0 (annuity-due)
PV_due <- 500 * (1 + ((1 - V^4) / i))
PV_due         # ~ 2317.50
```

## d) Effective annual rate from 21% nominal, compounded monthly

```{r p2d}
#| label: p2d
r <- 0.21
m <- 12
i_a <- (1 + r/m)^m - 1
i_a
i_a * 100   # in percent
```

## e) Future value of an annuity-due: \$1,000/yr, i = 3%, n = 5

```{r p2e}
#| label: p2e
P <- 1000
i <- 0.03
n <- 5
FV <- (1 + i) * P * (((1 + i)^n - 1) / i)
FV   # ~ 5468.41
```

# Problem 3: Fund Interest Rates & Capital Redemption Policy

|                                |                |
|--------------------------------|----------------|
| Fund at first January          | 1,25,00,000.00 |
| Premium income during the year | 10,00,000.00   |
| Claims                         | 2,50,000.00    |
| Management Expenses            | 1,25,000.00    |
| Gross interest                 | 12,50,000.00   |
| Income tax                     | 1,25,000.00    |

## a) Gross and net rate of interest earned by an insurance fund

```{r p3a}
#| label: p3a
A  <- 12500000    # Fund at start of year
PI <- 1000000     # Premium income
GI <- 1250000     # Gross interest earned
IT <- 125000      # Income tax
C  <- 250000      # Claims
E  <- 125000      # Management expenses

I <- GI - IT       # net interest
B <- A + PI + I - C - E
B   # 14,250,000

i_net <- (2 * I) / (A + B - I)
i_net; i_net * 100      # ~ 8.78%

I1 <- GI
i_gross <- (2 * I1) / (A + B - I)
i_gross; i_gross * 100  # ~ 9.76%
```

## b) Capital redemption policy: Tk. 10,000 at end of 10 years, i = 10%

```{r p3b}
#| label: p3b
i <- 0.10
n <- 10
V <- 1 / (1 + i)
K <- 10000

# (i) Annual premium
AC <- (V^n) * K
a_n     <- (1 - V^n) / i
a_n_due <- (1 + i) * a_n
P <- AC / a_n_due
P   # ~ 570.41

# Policy value at end of year 5
n2 <- 5
S_n     <- ((1 + i)^n2 - 1) / i
S_n_due <- (1 + i) * S_n
PValue  <- P * S_n_due
PValue  # ~ 3830.67

# (ii) Paid-up policy value at end of year 5
Paidvalue <- PValue / (V^(n - n2))
Paidvalue # ~ 6169.33
```

# Problem 4: Survival Analysis on Churn Data

## a) Load data & describe characteristics

```{r p4a}
#| label: p4a
data <- read.csv(file.choose())
dim(data)
head(data)
names(data)
str(data)
# The dataset has 21 variables (categorical + numerical), including
# State, Account.length, Area.code, Phone, Int.l.Plan, VMail.Plan,
# VMail.Message, Day/Eve/Night/Intl usage & charge fields,
# CustServ.Calls, and Churn (the outcome).
```

## b) Recode Churn as 0/1

```{r p4b}
#| label: p4b
dat <- data[, c("Account.length", "Churn")]
dat$Churn <- ifelse(dat$Churn == "TRUE", 1, 0)
dat$Churn
```

## c) Fit and plot the survival function

```{r p4c}
#| label: p4c
survdat <- Surv(time = dat$Account.length, event = dat$Churn)
fit <- survfit(survdat ~ 1, se = TRUE)
plot(fit, main = "Survival Function", xlab = "Time (Account length)",
     ylab = "Survival Probability")
fit
# Median time to churn ~ 201
```

## d) Fit Cox Proportional Hazards model

```{r p4d}
#| label: p4d
# Fixed: "<- " must have no space (original "< -" is a comparison, not
# assignment), and `data =` must reference the full data frame with
# Churn recoded, not an undefined object.
fitt <- coxph(
  Surv(Account.length, Churn) ~ State + Int.l.Plan + VMail.Plan + Day.Mins,
  data = within(data, Churn <- ifelse(Churn == "TRUE", 1, 0))
)
summary(fitt)
```

## e) Overall comment

Median time to churn is \~201 days. State, Int.l.Plan, VMail.Plan, and
Day.Mins are significant predictors. `Int.l.Plan` and higher `Day.Mins`
carry hazard ratios \> 1 (higher churn risk); `VMail.Plan` and longer
`Account.length` carry hazard ratios \< 1 (protective, lower churn
risk).

# Problem 5: Multiple Decrement Table

## 1-2) Assign probabilities and build the table

```{r p5-1-2}
#| label: p5-1-2
n <- 100
age <- 0:n
q1 <- rep(0.01, n + 1)   # death
q2 <- rep(0.02, n + 1)   # withdrawal
q3 <- rep(0.03, n + 1)   # disability

multi_table <- data.frame(age = age, q1 = q1, q2 = q2, q3 = q3)
multi_table
```

## 3-4) Total decrement rate and total survival probability

```{r p5-3-4}
#| label: p5-3-4
multi_table$total_q <- multi_table$q1 + multi_table$q2 + multi_table$q3
multi_table$p_tau   <- 1 - multi_table$total_q
```

## 5) Number of lives ($l_\tau$)

```{r p5-5}
#| label: p5-5
l0 <- 100000   # fixed: must be a variable name, not "10 <- 100000"
multi_table$l_tau <- 0
multi_table$l_tau[1] <- l0
for (i in 2:nrow(multi_table)) {
  multi_table$l_tau[i] <- multi_table$l_tau[i - 1] * multi_table$p_tau[i - 1]
}
multi_table$l_tau
```

## 6-8) Decrements by cause

```{r p5-6-8}
#| label: p5-6-8
multi_table$d1 <- multi_table$l_tau * multi_table$q1  # death
multi_table$d2 <- multi_table$l_tau * multi_table$q2  # withdrawal
multi_table$d3 <- multi_table$l_tau * multi_table$q3  # disability
multi_table
```

## 9) Decrements at age 15

```{r p5-9}
#| label: p5-9
multi_table[multi_table$age == 15, ]
# d1 (death)      ~ 395.29
# d2 (withdrawal) ~ 790.58
# d3 (disability) ~ 1185.88
```

## 10) Overall comment

Since q1, q2, q3 are constant, $l_\tau$ declines geometrically:
$l_\tau(x) = 100000 \times 0.94^x$. By age 15 about 60.5% of lives
remain (\~39,529); by age 100 only \~205 remain. Disability removes the
most lives at every age (highest rate, 0.03), followed by withdrawal
(0.02), then death (0.01) — consistently in roughly a 1:2:3 ratio.

# Problem 6: Joint-Life and Last-Survivor Status

Lives: $x$ (age 65), $y$ (age 62); constant mortality $q_x = 0.005$,
$q_y = 0.004$; interest $i = 5\%$; projected for $t = 0$ to $40$ years.

## 1) Mortality, survival probabilities, and discounting setup

```{r p6-1}
#| label: p6-1
age_x <- 65; age_y <- 62
qx <- 0.005; qy <- 0.004
i <- 0.05
v <- 1 / (1 + i)
n <- 40
t <- 0:n

px <- 1 - qx
py <- 1 - qy
tpx <- px^t
tpy <- py^t
```

## 2) Joint survival probability

```{r p6-2}
#| label: p6-2
tpxy <- tpx * tpy
```

## 3) Last-survivor probability

```{r p6-3}
#| label: p6-3
tp_last <- tpx + tpy - tpxy
```

## 4) Combined table at selected durations

```{r p6-4}
#| label: p6-4
select_t <- c(0, 1, 5, 10, 15, 20, 25, 30, 40)
cat(sprintf("%-6s  %-12s  %-12s  %-12s  %-14s\n",
    "Year", "tpx", "tpy", "tpxy", "tp(last surv)"))
for (k in select_t) {
  cat(sprintf("%-6d  %-12.6f  %-12.6f  %-12.6f  %-14.6f\n",
      k, tpx[k + 1], tpy[k + 1], tpxy[k + 1], tp_last[k + 1]))
}
```

## 5) Joint-life annuity-due

```{r p6-5}
#| label: p6-5
axy <- sum(tpxy * v^t)
axy   # ~ 16.1388
```

## 6) Last-survivor annuity-due

```{r p6-6}
#| label: p6-6
a_last <- sum(tp_last * v^t)
a_last   # ~ 18.0618

# Cross-check: ax + ay - axy = a_last
ax <- sum(tpx * v^t)
ay <- sum(tpy * v^t)
ax + ay - axy
```

::: callout-note
`tpxy` declines faster than either individual survival probability
because it requires *both* lives to survive; `tp_last` stays close to 1
much longer because only one life needs to survive. This is why the
last-survivor annuity (18.06) is worth more than the joint-life annuity
(16.14) — a joint-life benefit stops at the first death, while a
last-survivor benefit continues until the second death.
:::
